---
title: "BIOSCI738"
subtitle: ""
author: "Universty of Auckland"
institute: "Find course [bluepages here](https://stats-uoa.github.io/BIOSCI738/)"
date: ""
output:
  xaringan::moon_reader:
   yolo: 
     times: 1
     img: figs/yolo_img.jpg
   css: xaringan-themer.css
   lib_dir: libs
   nature:
     highlightStyle: github
     highlightLines: true
     countIncrementalSlides: false
     slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
        
---
class: inverse

# Assessment

 + **Workshops** 5% Total (10 to complete, worth 0.5% each) Students must attend at least 7 out of the 10 workshops)
 + **Assignments** 55% Total (5 to complete worth 10%) Each assignment will also contain a peer review aspect (worth 1%) students must complete the peer review coursework component of the assignments to be awarded their associated peer reviewed grade
 + **Final Exam** 40% Total
 + Throughout the course there will also be the options to complete additional assignments/tasks for extra credit (capped at 5% in total)

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_solarized_light()
```

<style type="text/css">
.remark-slide-content {
    font-size: 20px;
    padding: 1em 4em 1em 4em;
}
</style>


---
class: inverse, center, middle

## `R` and `RStudio` catchup
### Week 1

