## `R` and `RStudio`

Throughout this course we will be programming using `R`. `R` is a **free** **open-source** statistical programming language created by Ross Ihaka (of [Ngati Kahungunu, Rangitane and Ngati Pakeha](https://www.historyofdatascience.com/ross-ihaka-the-open-hearted-data-hero/) descent) and Robert Gentleman, here at UoA in the Department of Statistics! It is widely used across many scientific fields for data analysis, visualization, and statistical modeling. Proficiency in `R` will enable you to wrangle and explore datasets, conduct statistical analyses, and create visualizations to communicate your findings. These are all essential tools required in any scientific discipline.


<div class="alert alert-warning">
  <strong>TASK</strong> Go to the Statistics Department display in the science building foyer (building 302) to see a CD-Rom with the **first** ever version of `R`.
</div>

`RStudio` is an integrated development environment (IDE) for the `R` programming language. It serves as a user-friendly workspace, offering additional tools for coding, data visualization, and project management. `RStudio` simplifies the process of learning `R` by providing a structured interface with many built-in tools to help organize workflow, fostering a systematic approach to data analysis and research. 

<div class="alert alert-warning">
  <strong>TASK</strong> Research the meaning of open-source software and briefly outline the pros and cons of this in the context of statistical analysis.
</div>

### Recap: `R` terminology


| Term                | Description                                                                                                     |
|---------------------|-----------------------------------------------------------------------------------------------------------------|
| `Script`            | A file containing a series of `R` commands and code that can be executed sequentially.                           |
| `Source`            | To execute the entire content of an `R` script, often done using the "Source" button in `RStudio`.                   |
| `Running Code`      | The process of executing `R` commands or scripts to perform specific tasks and obtain results.                    |
| `Console`           | The interactive interface in `RStudio` where `R` commands can be entered and executed line by line.                |
| `Commenting`        | Adding comments to the code using the `#` symbol to provide explanations or annotations. Comments are ignored during code execution. |
| `Assignment Operator`| The symbol `<-` or `=` used to assign values to variables in `R`.                                                 |
| `Variable`          | A named storage location for data in `R`, which can hold different types of values.                                |
| `Data Type`         | The classification of data into different types, such as numeric, character, logical, etc.                       |
| `Object`            | A data structure that holds a specific type of data. Objects are used to store and manipulate data, and they can take various forms depending on the type of information being represented. |
| `Logical Operator`  | Symbols like `==`, `!=`, `<`, `>`, `<=`, and `>=` used for logical comparisons in conditional statements.      |
| `Function`          | A block of reusable code that performs a specific task when called (e.g., `mean(c(3, 4))`.                                             |
| `Argument`          | The input values that are passed to the function when it is called.                                              |
| `Error Handling`    | The process of anticipating, detecting, and handling errors that may occur during code execution.                |
| `Debugging`         | The process of identifying and fixing errors or bugs in the code.                                                |
| `Workspace`         | The current working environment in `R`, which includes loaded data, variables, and functions.                       |

### Which `R` syntax?

This is not a comprehensive list of `R` syntax, but a call to develop your own coding style and use whatever you are most comfortable with. Throughout this course I will likely use a mix of syntax and functions to carry out the same operations. This is (somewhat) by design, and is intended to expose you to a strength^[some might say] of `R`, which is that there are multiple ways of doing the same thing. As long as it works and is reproducible then whatever works for you, although of course there are some recommended practices, see [Good coding practice] .

**Example `<-` vs `=`**

Both `<-` and `=` are assignment operators. It is mainly a personal choice which you use^[[Ross Ihaka has been know to express his preference for `=` simply due to requiring *less typing*](https://www.youtube.com/watch?v=88TftllIjaY)]. However, there is a difference depending on which environment you evaluate the assigning. In the top level there is no difference; however, `=` does not act as an assignment operator in a function call, whereas `<-` does.

<div class="alert alert-warning">
  <strong>TASK</strong> Both lines of code below give you a two row matrix, can you work out what the difference is between them (HINT look at what is created in your environment after each line)?
  
```{r, eval = FALSE}
matrix("BIOSCI78", nrow = 2)
matrix("BIOSCI78", nrow <- 2)
```
</div>

**Example `%>%` vs `|>`**

Recall the `tidyverse` (specifically the `magrittr`) pipe operator `%>%`, which allows us to combine multiple operations in `R` into a single sequential chain of actions. There is also a base pipe `|>` which acts similarly, but not exactly. Essentially, `|>`  passes the left-hand side as the **first** argument in the right-hand side call. This is subtly different from `%>%` as it cannot pass the left-had side onto multiple arguments. The `%>%` operator, however, does allow for this and you can also change the placement of the left-hand object with a `.` placeholder^[although since v.4.2.0 the base pipe does now allow for a names placeholder `_`].

<div class="alert alert-warning">
  <strong>TASK</strong> Given `df <- data.frame(x = 1:10, group = rep(c("A", "B")))` are all the following operations equivalent?
  
```{r, eval = FALSE}
split(df, df$group)
df %>% split(.$group)
df |> split(df$group)
```
</div>

## Error handling and debugging

Sometimes rather than doing what you expect it to `R` will return an Error message to you via the *Console* prefaced with **Error in...** followed by text that will try to explain what went wrong. This, generally, means something has gone wrong, so what do you do?

 1. **Read it**! THE MESSAGES ARE WRITTEN IN PLAIN ENGLISH (MOSTLY)
 2. **DO NOT** continue running bits of code hoping the issue will go away. IT WILL NOT.
 3. Try and work out what it means and fix it (some tips below).

Interpreting `R` errors is an invaluable skill! The error messages are designed to be clear and informative. They aim to tell might be going wrong in the code. These messages typically contain information about the type of error hinting at how you might fix it. For example, if there is a typo in a function or variable name, `R` will produce an error like `"Error: object 'variable_name' not found.`" This suggests that `R` couldn't find the specified object `variable_name`.

Understanding error messages involves paying attention to the details, such as the error type (e.g., syntax error, object not found, etc.), and the specific line number where the error occurred! You should always run your code line-by-line, especially if you are new to programming. This means that the execution of each line of code is done in in isolation, providing immediate feedback and pinpointing the exact location where an error occurs. If the meaning or solution to an error isn't immediately obvious then make use of the documentation (RTFM) and even technology. Online platforms like [Stack Overflow](https://stackoverflow.com/) or [RStudio Community](https://community.rstudio.com/) and tools like [ChatGPT](https://chat.openai.com/) can often be your friend! It is very unlikely that you are the first one to have encountered the error and some kind soul with have posted a solution online (which the AI bot will have scraped...). However, it's crucial to approach online answers carefully. You need to first understand the context of your error and use your knowledge about your issue to figure out if the solution is applicable (this gets easier with experience).


It is always best to try debugging yourself before blindly following a stranger's advice. Debugging is a crucial aspect of `R` programming and the process itself helps solidify your understanding. There are several widely used methods that help identify and resolve issues in your code, two are outlined below.

 + [**Print debugging**](https://en.wikipedia.org/wiki/Debugging) involves actively printing out objects and asking the software to display variable values or check the flow of execution. For example, if you get this error message `"Error: object 'variable_name' not found.` I might ask `R` to print out what objects do exist in my workspace, it could be that a previous typo means that the object `variable_nam` exists.
 + [**Rubber duck debugging**](https://en.wikipedia.org/wiki/Rubber_duck_debugging) is a useful debugging strategy where you explain your code or problem out loud (as if explaining it to a rubber duck). Honestly it works! The process helps clarify your thoughts and identify issues even before seeking external help.

The table below summaries some common `R` issues and errors and their solutions.

| Error/Issue                               | Description                                                                                           | Solution                                                                                                 |
|--------------------------------------------|-------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------|
| **Syntax Error**                           | Incorrect use of `R` syntax, such as missing or mismatched parentheses, braces, or quotes.              | Carefully review the code, check for missing or mismatched symbols, and ensure proper syntax is used.     |
| **Object Not Found**                       | Attempting to use a variable or function that hasn't been defined or loaded into the workspace.       | Check for typos in variable or function names, ensure the object is created or loaded, and use correct names. |
| **Package Not Installed/Loaded**           | Trying to use a function from a package that is not installed or loaded into the environment.         | Install the required package using `install.packages("package_name")`, and load it using `library(package_name)`. |
| **Undefined Function/Variable**            | Using a function or variable that hasn't been defined or is out of scope.                             | Define the function or variable, or check its scope within the code.                                       |
| **Data Type Mismatch**                     | Operating on data of incompatible types, such as performing arithmetic on character data.            | Ensure that data types are compatible, and consider using functions like `as.numeric()` or `as.character()` for conversions. |
| **Misuse of Assignment Operator**          | Using the wrong assignment operator (`=` instead of `<-` or vice versa).                              | Be consistent with the assignment operator, and use `<-` for assignment in most cases.                   |
| **Missing Data**                           | Dealing with missing values in the data, leading to errors in calculations or visualizations.        | Handle missing data appropriately, using functions like `na.omit()` or `complete.cases()`.               |
| **Index Out of Bounds**                    | Attempting to access an element in a vector or data frame using an index that is out of range.        | Check the length of the vector or data frame and ensure the index is within bounds.                       |
| **Failure to Load a File**                 | Issues with loading a data file using functions like `read.csv()` or `read.table()`.                 | Check the file path, file format, and encoding. Confirm that the file exists in the specified location.   |
| **Incorrect Function Arguments**           | Providing incorrect arguments or parameters to a function, leading to errors.                         | Refer to the function's documentation to understand the correct parameters and ensure proper usage.      |


Sometimes your computer will return a warning messages to you prefaced "Warning:". These can sometimes be ignored as they may not affect us. However, READ THE MESSAGE and decide for yourself. Occasionally, also your computer will write you a friendly message, just keeping you up-to date with what it's doing, again don't ignore these they might be telling you something useful!


<div class="alert alert-warning">
  <strong>TASK</strong> Keep a bug diary! Each time you get an error message, see if you can solve it and keep a diary of your attempts and solution.
  
</div>
