## Inference for a linear model

After all that what do estimated parameters mean? Assuming our model assumptions are met we can draw inference based on the estimated parameter values.

### Point prediction

Using the `slm_sp` model we can make a point prediction for the expected bill depth (mm) for Gentoo penguins with a bill length of 50mm.

Recall the model equation is


```{r, echo = FALSE}
equatiomatic::extract_eq(slm_sp,wrap = TRUE, terms_per_line = 2)
```

The fitted equation is given as

```{r, echo = FALSE}
equatiomatic::extract_eq(slm_sp,wrap = TRUE, terms_per_line = 2, use_coefs = TRUE, coef_digits = 2)
```

We can then simply substitute in the values:


$$\widehat{\text{bill depth}} = \hat{\alpha} + \hat{\beta_1}*50 + \hat{\beta_3}*1$$
$$\downarrow$$

$$\widehat{\text{bill depth}} = 10.56 + 0.20*50 - 5.10*1$$
$$\downarrow$$

$$15.47\text{mm}$$

Rather than *by hand* we can do this easily in `R`

```{r,echo = TRUE}
## create new data frame with data we want to predict to
## the names have to match those in our original data frame
newdata <- data.frame(species = "Gentoo",bill_length_mm = 50)
## use predict() function
predict(slm_sp, newdata = newdata) ## more accurate than our by hand version!
```

What does this look like on a plot?

```{r, echo = FALSE}
## calculate predicted values
penguins$pred_vals <- predict(slm_sp)
pred_data <- data.frame(x = c(-Inf,50), y = c(predict(slm_sp, newdata = newdata),-Inf))
ggplot(data = penguins, aes(y = bill_depth_mm, x = bill_length_mm, color = species)) +
  geom_point() + ylab("Bill depth (mm)") +
  xlab("Bill length (mm)") + theme_classic()  +
  geom_line(aes(y = pred_vals)) + 
  geom_segment(data = pred_data, aes(x = x[1], y = y[1], xend = x[2], yend = y[1]), 
               inherit.aes = FALSE) + 
  geom_segment(data = pred_data, aes(x = x[2], y = y[1], xend = x[2], yend = y[2]), 
               inherit.aes = FALSE) 
```


### Confidence intervals for parameters

For the *chosen* `slm_sp` model we can get these simply by using `confint()`. By default the 95% intervals are returned:


```{r, echo = TRUE}
cis <- confint(slm_sp)
cis
```

So this tells us that For every 1mm increase in bill length we estimate the expected bill depth to increases between `r round(cis[2,1],3)` and `r round(cis[2,2],3)` mm

We estimate that the expected bill depth of a Chinstrap penguin is between
`r -1*round(cis[3,2],1)` and `r -1*round(cis[3,1],1)` mm shallower than the Adelie penguin

### Marginal predictions

One other thing we may want to do is give marginal precictions which show the marginal effect of a predictor, holding other variables constant.

The `ggpredict()` function from the `ggeffects` package can be used for this:

```{r}
ggpredict(slm_sp, terms = "species")
```

We can even plot the marginal effects:

```{r}
ggpredict(slm_sp, terms = "species") %>% 
  ggplot(aes(x = x, y = predicted, label = round(predicted, 3))) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = .5) +
  geom_text(hjust = -.2) +
  ylab( "Predicted bill depth (mm)") + xlab("") +
  ggtitle("Fitted means for each species with 95% CIs", subtitle = "Adjusted for bill_length_mm = 43.99") +
  theme_classic()

```
