## Permutation tests

The basic approach to permutation tests is straightforward:

 1. Choose a statistic to measure the effect in question (e.g., differences between group means)
 2. Calculate that test statistic on the observed data. Note this metric can be **anything** you wish
 3. Construct the sampling distribution that this statistic would have if the effect were not present in the population (i.e.,
 the distribution under the *Null* hypothesis, $H_0$): For chosen number of times
    + shuffle the data labels
    + calculate the test statistic for the reshuffled data and retain
 4. Find the location of your observed statistic in the sampling distribution. The location of observed statistic in sampling distribution is informative:
   + if in the main body of the distribution then the observed statistic could easily have occurred by chance
   + if in the tail of the distribution then the observed statistic would rarely occur by chance and there is evidence that something other than chance is operating.
 5. Calculate the proportion of times your reshuffled statistics equal or exceed the observed. This *p-value* is the probability that we observe a statistic at least as “extreme” as the one we observed **State** the strength of evidence against the null on the basis of this **probability**.

### Significance testing using permutation (*randomisation*) tests

#### Permutation Test on Two Independent Samples

##### Pāua shell lengths

Remember the Pāua data from [Chapter 1](https://stats-uoa.github.io/BIOSCI738/r-rstudio-and-git.html#exploratory-data-analysis) One question we may want to ask is if on average the shell length differs between Species? 

**Scientific question**: Are the shell lengths of shells the same in both species?
**Null hypothesis**: The distribution of shell lengths in *Haliotis iris* the same as in *Haliotis australis*
**Test statistic**: Difference of sample means


```{r violen}
means <- paua %>% group_by(Species) %>% summarise(means = mean(Length))
ggplot(paua,aes(x = Species, y = Length)) + 
  geom_violin() +
  geom_point(alpha = 0.4) +
  ylab("Length (cms)") + xlab("") +
  theme_classic() +
  geom_point(data = means, aes(x = Species, y = means, color = Species), size = 2) +
  geom_hline(data = means, aes(yintercept = means, color = Species), lty = 2, alpha = 0.5) +
  theme(legend.position = "none") +
  geom_text(data = means, aes(x = Species, y = means + 0.3, label = paste0("Species averege = ",round(means,3)), color = Species))
  
ggplot(paua,aes(x = Length, fill = Species)) + 
  geom_histogram(position = "identity", alpha = 0.3) +
  xlab("Length (cms)") + ylab("") +
  theme_classic()

```

But because the data are skewed and we've likely got non-constant variances we may be better off adopting a randomization test, rather than a parametric t-test

```{r}
## observed differences in means
diff_in_means <- (paua %>% group_by(Species) %>%
                    summarise(mean = mean(Length)) %>% 
                    summarise(diff = diff(mean)))$diff
diff_in_means
## Number of times I want to randomise
nreps <- 1000   
## initialize empty array to hold results
randomisation_difference_mean <- numeric(nreps)
set.seed(1234) ## *****Remove this line for actual analyses*****
## This means that each run with produce the same results and
## agree with the printout that I show.

for (i in 1:nreps) {
  ## the observations
  data <- data.frame(value = paua$Length)
  ##  randomise labels
  data$random_labels <-sample(paua$Species, replace = FALSE)
  ## randomised differences in mean
  randomisation_difference_mean[i] <- (data %>% group_by(random_labels) %>% summarise(mean = mean(value)) %>% summarise(diff = diff(mean)))$diff
}
## results
results <- data.frame(randomisation_difference_mean = randomisation_difference_mean)
```

```{r pval}
## How many randomised differences in means are as least as extreme as the one we observed
## absolute value as dealing with two tailed
n_exceed <- sum(abs(results$randomisation_difference_mean) >= abs(diff_in_means))
n_exceed
## proportion
n_exceed/nreps
```


```{r hist}
ggplot(results, aes(x = randomisation_difference_mean)) +
  geom_histogram() +
  theme_classic() + ylab("") + xlab("Differences between randomised group means") +
  geom_vline(xintercept = diff_in_means, col = "cyan4", size = 1,alpha = 0.6) +
  annotate(geom = 'text', label = "Observed difference between means" , 
           x = -Inf, y = Inf, hjust = 0, vjust = 1.5, color = "cyan4")
  

```

How would the parametric t-test have served?

```{r t}
t.test(Length ~ Species, data = paua)
```

Not too different after all

##### Jackal mandible lengths

```{r}
## Mandible lengths (mm) for  golden jackals (Canis aureus) of each sex from the British Museum
jackal <- data.frame(mandible_length_mm = c(120, 107, 110, 116, 114, 111, 113, 117, 114, 112,
                                            110, 111, 107, 108, 110, 105, 107, 106, 111, 111),
                     sex = rep(c("Male","Female"), each = 10))
```


**Scientific question**: Are the jaw lengths of jackals the same in both sexes?
**Null hypothesis**: The distribution of jaw lengths in male jackals the same as in in females
**Test statistic**: Difference of sample means

```{r, echo = FALSE}
ggplot(jackal, aes(x = sex, y = mandible_length_mm)) +
  geom_violin() +
  geom_jitter(alpha = 0.5) +
  xlab("Biological Sex") +
  ylab("Mandible length (mm)") +
  ggtitle("Mandible lengths for golden jackals (Canis aureus) ")
```

Rather than a *for loop* let's try this another way.

```{r}
## observed statistic
jackal_mean_diff <- (jackal %>%
  group_by(sex) %>%
  summarise(mean = mean(mandible_length_mm)) %>% 
    summarise(diff = diff(mean)))$diff
## Generate all possible combinations
## This time we're doing ALL possble ones
## rather than a rendom 1000
combinations <- combn(20,10)
## Do the permutations
permtest_combinations <- apply(combinations, 2, function(x)
  mean(jackal$mandible_length_mm[x]) - mean(jackal$mandible_length_mm[-x]))
## Full Permutation test p.value
length(permtest_combinations[abs(permtest_combinations) >= jackal_mean_diff]) / choose(20,10)
## Now let's use 10000 random permutations, sample without replacement
## set up matrix
random_perm <- apply(matrix(0, nrow = 10000, ncol = 1), 1, function(x) sample(20))
random_mean_diff <- apply(random_perm, 2, function(x){
  z <- jackal$mandible_length_mm[x]
  mean(z[jackal$sex == "Male"]) - mean(z[jackal$sex == "Female"])
})
random_p.value <- length(random_mean_diff [abs(random_mean_diff) >= jackal_mean_diff]) / 10000 ## note the abs()
random_p.value
## Now what about a t-test (two-sample)
t.test(mandible_length_mm ~ sex, data = jackal)$p.value
```


**NOTE: We can extend the randomization test to make inference about any sample statistic (not just the mean)**

