## Modler v. designer: the maths

### A completely randomised design (CRD)

Last Christmas I was given a gift set of three types of coffee beans. I want to know which makes the darkest coffee; to do this I measure the opacity after the coffee is made. To work this out I set up a completely randomized experiment where each of 12 cups of are made and the type of coffee used randomly assigned to each cup. Below is the experiment plan I used and and the meaured outcome.

```{r, echo = FALSE}
set.seed(5469) 
type <- factor(rep(paste("Type", 1:3), each = 4))
df <- data.frame(treatment = type, opacity = round(c(rnorm(4, 5), rnorm(4, 7), rnorm(4, 9)), 1))
df <- df[sample(1:12), ]
df$cup <- 1:12; df <- df[, c(3, 1, 2)]
df
```

#### As a linear model

As we've seen in the previous module that we can write a linear model with a single explanatory variable as 

$$Y_i = \alpha + \beta_1x_i + \epsilon_i$$

When dealing with factor variables we use dummy variables and can write the above as
 
 
 $$Y_{ik} = \alpha + \tau_k + \epsilon_{ik}$$ where $\tau_k$ is called an *effect* and represents the difference between the overall average, $\alpha$, and the average at the $k_{th}$ treatment level. The errors $\epsilon_{ik}$ are again assumed to be normally distributed and independent due to the randomisation (i.e., $\epsilon_{ik} \sim N(0, \sigma^2)$.


Or you might think of the model as

$$Y_{ik} = \mu_k + \epsilon_{ik}$$

where $Y_{ik}$ is the response (i.e., observed coffee opacity) for the $i^{th}$ experimental unit (i.e., coffee cup) subjected to the $k^{th}$ level of the treatment factor (i.e., coffee type).  Here $\mu_k$ are the different (cell) means for each level of the treatment factor. See below for an illustration of this for three factor treatment levels (as in the coffee example above).

```{r, echo = FALSE, fig.height=5}
require(tidyverse)
ggplot(data.frame(x = c(-4, 17)), aes(x)) + 
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1)) +
  stat_function(fun = dnorm, args = list(mean = 5, sd = 1))+
  stat_function(fun = dnorm, args = list(mean = 12, sd = 1)) +
  theme_classic() +
  scale_x_continuous(breaks = c(0, 5, 12), labels = c(expression(mu[1]), expression(mu[2]), expression(mu[3]))) +
  xlab("") + ylab("") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```




