# Ordination

## Multidimensional Scaling (MDS), AKA *Principal Coordinate Analysis (PCO)*

Consider data which are not represented as points in a feature space: 

  + Where we are only provided with (dis)similarity matrices between objects, e.g. chemical compounds, images, trees, or other complex objects
  + WHere there is no obvious coordinates in (continuous) n-dimensional space .

### Examples in `R`


**Distances (in km) between North Island cities**

File `north_island_distances.csv` can be found on CANVAS

```{r, echo = FALSE}
library(tidyverse)
ni <- read_csv("../data/north_islands_distances.csv")[,-1]
```

```{r, eval = FALSE}
library(tidyverse)
ni <- read_csv("../data/north_islands_distances.csv")[1,-]
```

```{r}
library(pheatmap)
pheatmap(ni, cluster_rows = TRUE,
         treeheight_row = 2, treeheight_col = 2,
         fontsize_row = 12, fontsize_col = 12,
         cellwidth = 26, cellheight = 26)


```

```{r}
mds <- cmdscale(ni, eig = TRUE)
mds
```


**Eckmans colour perception (1954)**

Data may from objects for which we have similarities but no underlying (geometric) space. Here the goal is to understand the underlying dimensionality of colour perception.

  + Similarities for 14 colours, with wavelengths from 434 to 674nm based on rating by 31 subjects
  + Each pair of colours was rated on a 5-point scale:
      + 0 = no similarity up to 4 = identical.
  + After averaging over 31 raters the similarities were divided by 4 such that they are within the unit interval.

Data available on CANVAS

```{r, echo = FALSE}
library(tidyverse)
ekman <- read_csv("../data/ekman.csv")
ekman = 1 - ekman - diag(1, ncol(ekman))
ekman[1:5, 1:5]
```

```{r, eval = FALSE}
library(tidyverse)
ekman <- read_csv("ekman.csv")
```


```{r}
ekman.mds <- cmdscale(ekman, eig = TRUE)
ekman.mds
round(ekman.mds$eig,2)
```

**Distances (in km) between 21 cities in Europe**

```{r,eval = TRUE}
library(ggfortify)
## Plotting Multidimensional Scaling (for interest)
## stats::cmdscale performs Classical MDS
data("eurodist") ## road distances (in km) between 21 cities in Europe.
autoplot(eurodist)
## Plotting Classical (Metric) Multidimensional Scaling
autoplot(cmdscale(eurodist, eig = TRUE))
autoplot(cmdscale(eurodist, eig = TRUE), label = TRUE, shape = FALSE,
         label.size = 3)
## Plotting Non-metric Multidimensional Scaling
## MASS::isoMDS and MASS::sammon perform Non-metric MDS
library(MASS)
autoplot(sammon(eurodist))
autoplot(sammon(eurodist), shape = FALSE, label = TRUE,label.size = 3)
## Have a go at interpreting these plots based on the geography of the cities :-)
```

## Non-metric Multidimensional Scaling

Multidimensional scaling aims to minimize the difference between the squared distances $D^2$ from the distance matrix $D$, and the squared distances between the points with their new coordinates. Unfortunately, this objective tends to be sensitive to outliers: one single data point with large distances to everyone else can dominate, and thus skew, the whole analysis.


### Examples in `R`

```{r}
library(vegan)
nMDS.2 <- replicate(100, metaMDS(ekman, k = 2, autotransform = FALSE)) 
quantile(unlist(nMDS.2["stress", ]))

```


```{r}
stressplot(metaMDS(ekman, k = 2, autotransform = FALSE), pch = 20, cex = 2)

```


  
## Correspondence Analysis (CA)

CA is a special case of metric MDS where the distance measure is the chi-square distance. It is conceptually similar to principal component analysis but where the data are categorical, counts, rather than continuous. CA is traditionally applied to contingency tables where rows and columns are treated equivalently; it decomposes the chi-square statistic associated with this table into orthogonal factors.

```{r}
HairEyeColor
```

```{r}
library(ade4)

```


## Summary

**Multivariate data**
Measured on a variety of scales, taking a variety of forms. If data are – approximately – continuous measurements can use PCA to produce a lower dimensional
representation. For general data construct a matrix of distances
between sample points: use distance matrix to construct a geometric representation, distances between geometric points approximate distances from
distance matrix. For Count data Correspondence Analysis widely used.


PCO and metric MDS will usually provide almost identical answers if Kruskal’s STRESS is used

**Metric or non-metric?**
  + Metric has few advantages over Principal Coordinates Analysis (unless many negative eigenvalues)
  + Non-metric does better with fewer dimensions but can be more prone to sub-optimal solutions.
  
**How many dimensions?**
  + STRESS <10% is “good representation”
  + Scree diagram
  
